---
title: "Multi-set Intersection Analysis Using SuperExactTest"
author: "Minghui Wang, Yongzhong Zhao, Bin Zhang"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SuperExactTest User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8](inputenc)
---

```{r setup, include=FALSE}
library(knitr)
```

## Introduction

`SuperExactTest` provides statistical methods and visualization tools for analyzing intersections among multiple sets. This R package implements an 
exact probability framework for evaluating whether observed multi-set intersections are larger or smaller than expected by chance (Wang et al., 2015).

Unlike pairwise overlap tests, `SuperExactTest` computes the exact distribution of intersections involving an arbitrary number of sets using a 
forward algorithm whose computational complexity increases linearly with the number of sets. The package also includes scalable visualization 
methods for exploring intersection sizes and their statistical significance.

## Citation
If you use `SuperExactTest` in published work, please cite:

> Minghui Wang, Yongzhong Zhao, and Bin Zhang (2015) Efficient Test and Visualization of Multi-Set Intersections. *Scientific Reports* 5: 16923. DOI: [10.1038/srep16923] (https://doi.org/10.1038/srep16923)

## Case study

This vignette demonstrates the main functionality of `SuperExactTest` using expression quantitative trait loci (eQTLs) identified from four human brain regions.

eQTLs are genomic loci whose genetic variants influence gene expression. Because gene regulation varies across tissues, cell types, and developmental stages, 
comparing eQTLs across tissues can help identify shared and tissue-specific regulatory mechanisms.

The example dataset consists of genes with cis-eQTLs reported by Gibbs et al. (2010). These data were originally obtained from the NCBI eQTL Browser (now discontinued) 
and are included in `SuperExactTest` as a built-in example dataset.

Load the package and example data:

```{r}
library("SuperExactTest")
data("eqtls")
```
The object `cis.eqtls` is a named list containing four vectors of gene symbols, corresponding to the cerebellum (CB), frontal cortex (FC), temporal cortex (TC), and pons (PONS).
```{r}
str(cis.eqtls)
```
These cis-eQTL gene sets were identified from genome-wide expression profiling of 18,196 unique genes (Gibbs et al., 2010), which serves as the background population throughout this vignette.

The length of the *cis*-eQTL gene sets varies from 101 to 164:
```{r}
(length.gene.sets=sapply(cis.eqtls,length))
```
Assuming the *cis*-eQTL gene sets were independently and randomly sampled from the population of 18,196 unique genes profiled in the eQTL study,
the probability of the number of common genes shared by the four *cis*-eQTL gene sets can be computed using function `dpsets`
which implements the exact probability calculation of multi-set intersection developed in this study.


Before testing the observed overlap, we first calculate the expected intersection size under the null hypothesis that the four gene sets were independently sampled from the background population.
```{r}
total=18196
(num.expcted.overlap=total*do.call(prod,as.list(length.gene.sets/total)))
```
Due to the large background population gene size whereas small gene set sizes, the expected intersection size is close to 0.
It is obvious that the possible number of genes shared among the four *cis*-eQTL gene sets is from 0 to 101.
We can compute the probability density distribution of the possible intersection sizes using the `dpsets` function:
```{r}
(p = sapply(0:101, function(i) dpsets(i, length.gene.sets, n=total)))
```
In the function call of `dpsets`, the first argument is the intersection size, the second argument is a vector of the set sizes,
and option $n$ specifies the size of the background gene population from which the gene sets are collected.
As expected, the probability density is maximized at intersection size 0 and decreases as the intersection size increases.
When the intersection size is larger than 54, the p values become extremely low to 0 due to computational precision.
To recover the extremely low p values, we can use the `log.p` option to return the the p values in the logarithm space.
```{r}
(p.log = sapply(0:101, function(i) dpsets(i, length.gene.sets, n=total, log.p=TRUE)))
```

Next, let us compute the observed intersection among the four gene sets and then the corresponding fold enrichment (FE) by:

```{r}
common.genes=intersect(cis.eqtls[[1]], cis.eqtls[[2]], cis.eqtls[[3]],
 cis.eqtls[[4]])
(num.observed.overlap=length(common.genes))
(FE=num.observed.overlap/num.expcted.overlap)
```

Note that we extend the built-in `R` function `intersect` to make it capable of handling more than two input vectors.

The probability of observing exactly this intersection size is obtained with `dpsets()`.
```{r}
dpsets(num.observed.overlap, length.gene.sets, n=total, log.p=TRUE)
```
Again, we use the `log.p` option to recover the extremely low p value in the logarithm space.

The probability of observing 56 or more intersection genes can be calculated using the cumulative probability function `cpsets`:
```{r}
cpsets(num.observed.overlap-1, length.gene.sets, n=total, lower.tail=FALSE, log.p=TRUE)
```
Like function `dpsets`, `cpsets` takes similar arguments as input, with an additional argument `lower.tail` specifying which of the one-tail probabilities to be returned.
`num.observed.overlap - 1` because the probability is $P(X \leq x)$ when `lower.tail = TRUE` while $P(X > x)$ when `lower.tail = FALSE`.

The extremly low one-tail probability of the observed intersection size underlies highly significant overlap of common *cis* regulators across four brain regions in the present *cis*-eQTL dataset,
rejecting the null hypothesis that the *cis*-eQTL gene sets were independent random samples from the population of 18,196 unique genes profiled.

As a convenience, the `MSET()` function performs both the intersection calculation and the statistical test in a single step.
```{r}
fit = MSET(cis.eqtls, n=total, lower.tail=FALSE, log.p=TRUE)
fit
```

### Analyzing all possible intersections among four *cis*-eQTLs gene sets
In the above, we analyzed one specific intersection that is shared by all four gene sets. In practice, we may be also interested in the intersections among any combinations of the gene sets, such as overlap among any two or three gene sets in this case.

In many applications, all possible intersections—not only the intersection shared by every set—are of interest. The `supertest()` function enumerates all $2^n-1$ possible intersections for `n` input sets and performs the corresponding statistical tests automatically.
```{r}
res = supertest(cis.eqtls, n=total)
```

The returned variable `res` is an object of new `S3` class `"msets"`, which is a specifically designed list holding the analysis results and can be processed by generic functions including `plot` and `summary`.
For example, to visualize the analysis result, we can plot the analysis results in a circular layout as shown in the following figure:

```{r fig1, fig.width = 5, fig.height = 5, fig.cap = "A circular plot visualizing all possible intersections and the corresponding statistics amongst *cis*-eQTL gene sets. For convenience, the minimum p value is capped at 1.0E-320."}
plot(res, sort.by="size", margin=c(2,2,2,2), color.scale.pos=c(0.85,1), legend.pos=c(0.9,0.15))
```

The four tracks in the middle represent the four gene sets, with individual blocks showing "presence" (green) or "absence" (grey) of the gene sets in each intersection. To denote the four gene sets with different colors, set argument `color.on` to NULL or provide it with a vector of user-defined colors.
The height of the bars in the outer layer is proportional to the intersection sizes, as indicated by the numbers on the top of the bars. The color intensity of the bars represents the P value significance of the intersections.

As the option name suggests, `sort.by="size"` instructs the intersections to be sorted by size. Users are flexible to use different color schemes or sort intersections by P value, set, size or degree.
The complete control options can be obtained from the help documentation by command `help("plot.msets")`.
For example, we can visualize the results in a landscape (matrix) layout as shown in Figure which includes only intersections among 2 to 4 sets by option `degree = 2:4`:

```{r fig2, fig.width = 9, fig.height = 4, fig.cap = "A bar chart illustrating all possible intersections among *cis*-eQTL gene sets in a matrix layout."}
plot(res, Layout="landscape", degree=2:4, sort.by="size", margin=c(0.5,5,1,2))
```

The matrix of solid and empty circles at the bottom illustrates the "presence" (solid green) or "absence" (empty) 
of the gene sets in each intersection. The numbers to the right of the matrix are set sizes. The colored bars on the top of the matrix represent the intersection sizes with the color intensity showing the P value significance.


The generic `summary` function can be used to summarize the analysis results in details:
```{r}
summary(res)
```
The main output from the `summary` function is a data.frame, with each row representing an intersection and its corresponding test statistics.
Details about the `summary` function output are available from the help documentation:
```{r, eval=FALSE}
?summary.msets
```
To tabulate the `summary` result into a file, use command:
```{r, eval=FALSE}
write.csv(summary(res)$Table, file="summary.table.csv", row.names=FALSE)
```

# References
> Minghui Wang, Yongzhong Zhao, and Bin Zhang (2015) Efficient Test and Visualization of Multi-Set Intersections. *Scientific Reports* 5: 16923. DOI: [10.1038/srep16923] (https://doi.org/10.1038/srep16923)

> Gibbs et al. (2010) Abundant Quantitative Trait Loci Exist for DNA Methylation and Gene Expression in Human Brain. PLoS Genetics 6: e1000952.
